### 2019-09-28 paper-two/paper-two.tex
p2: brain functions started

\documentclass{article}

@@

I took my first Vipassana course in 2017 and since then I have been trying to wade through the research that has been published in this field, especially white papers dealing with the effects of a 10-day course on the brain. There is not much in terms of this research and the research that does exist usually uses a sample of experienced meditators and/or self-reports as a outcome. I have tried to write down what I experienced during my 10-day courses over the past year and tried to corroborate it with research that does exist regarding meditation and the brain.
Disclaimer: My primary field of research is biology but I am far removed from neuroscience. I have tried to simplify the work out there so as to better understand it. So, if there are any corrections to make or editing in terms of the content, please feel free to get in touch with me.
Before going any further, I will start by describing the different parts of the brain that have shown up in scientific literature as correlated to meditative practices.
Parts of the brain and their functions
Default Mode Network:
These are a set of structures in the mid-line of the brain (which is a evolutionarily more primitive area related to memory and emotion) which connects to structures in the cortex (which is evolutionarily the most recent part of the brain and contains the executive and higher order functions). It is the constellation of regions that light up when the brain is not engaged in any external/goal-oriented tasks (1). These regions are functionally connected and are involved in
a) The neurological basis of the self —
The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain. (Leech R, Sharp DJ (July 2013). “The role of the posterior cingulate cortex in cognition and disease”. Brain. 137 (Pt 1): 12–32.)
The PCC has been thought to mediate interactions between emotion and memory. It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes. It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).

@@

Somatosensory and Motor cortices
Another strange thing that starts happening is as your mind wandering slows down and you sit with your eyes closed, your sense of your self as a subjective agent in space and time changes, first just very slightly, as you close your eyes and meditate, it feels like you have shifted in space or turned in different directions (during these times the vmPFC and PCC which are a part of the DMN become less active), and then as the meditations get deeper and sometimes, the DMN quietened down completely for a very short duration and in addition to this, the insula is firing more rapidly, it could lead to other alternative networks of consciousness arising. It is as if losing the influence of one’s narrative leads to new insights that are usually kept from consciousness. Experiencing this even for a few seconds can lead to lasting changes even after this activity wears off. These experiences can be quite similar to being on certain kinds of drugs or even in some cases of ecstatic seizures where the abnormal activity of the anterior insula leads to heightened self awareness and feelings of bliss and lack of ambiguity. (Fabienne Picard, State of belief, subjective certainty and bliss as a product of cortical dysfunction, Cortex, Volume 49, Issue 9, 2013,
Pages 2494–2500ISSN 0010 9452, https://doi.org/10.1016/j.cortex.2013.01.006.)
So to conclude, I believe attending a ten day course literally did change the habit pattern of my mind at least for a while.
It quietens the Default Mode Network and causes the Salience Network and certain parts of the Central Executive Network to fire more rapidly (the parts that are aware of the sensations while quietening the reacting part of this network). This causes changes to a persons behavior at least for a short while and this is usually seen as soon as a person leaves the course. But repeated practice is essential for any sustained neuroplasticity since rewriting many years of habit formation requires more than a ten-day course. However, the ten day course is well designed to give the minimum observational change neccesary for people interested to keep practicing.
There have also been studies looking at the structural changes in the brain and functional changes especially to the default mode network and the salience network but these are usually measured after one has clocked over thousands of hours of practice. Therefore I am not delving into those right now.
How is this different from just concentrating on tasks hard? It is different because when concentrating on any task, your default mode network is quetened and the salience network and CEN are firing entering you into a state of flow but the reacting part of the brain is also working and working hard.
There are aspects to this technique that I have not understood or experienced that I have not delved into right now and may do once i know it better.
I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.
Reference
Andrews-Hanna, Jessica R. (2012–06–01). “The brain’s default network and its adaptive role in internal mentation”. The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC 3553600. PMID 21677128.
Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H. (January 2003). “Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”. Human Brain Mapping. 18 (1): 30–41.
Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan (2014–05–01). “The default network and self-generated thought: component processes, dynamic control, and clinical relevance”. Annals of the New York Academy of Sciences. 1316(1): 29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540.
Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
Isoda, M., \& Noritake, A. (2015). What makes the dorsomedial frontal cortex active during reading the mental states of others?. Neural basis of social learning, social deciding, and other-regarding preferences, 51.
Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
Menon, V; Toga, A (2015). Salience Network. Elsevier. pp. 597–611. ISBN 978–0–12–397316–0.
Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990). “The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”. Proceedings of the National Academy of Sciences of the United States of America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583.
Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002). “Dorsal anterior cingulate cortex: a role in reward-based decision making”. Proceedings of the National Academy of Sciences of the United States of America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID 11756669.
Sevinc G, Gurvit H, Spreng RN (July 2017). “Salience network engagement with the detection of morally laden information”. Social Cognitive and Affective Neuroscience. 12 (7): 1118–1127. doi:10.1093/scan/nsx035. PMID 28338944.
Jackson PL, Brunet E, Meltzoff AN, Decety J (2006). “Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”. Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID 16140345.
Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
Neuroreport. 11(7):1581–1585, May 15, 2000.
J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal, C. Morgado-Valle. The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
https://doi.org/10.1016/j.nrleng.2018.05.006.
REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
Emotion. Annals of the New York Academy of Sciences, 1121:
72–86. doi:10.1196/annals.1401.026

@@

>>>>

\documentclass[twocolumn]{article}

@@

I took my first Vipassana course in 2017. Vipassana meditation was so unlike anything
I had ever experienced before, after the course I was left extremely curious
about what exactly had happened to me during those ten days. For months afterward, I
spent my mornings and evenings wading through the research in the field of
meditation, with a specific focus on white papers dealing with the effects a
10-day Vipassana course has on the brains of participants. The research in this area
is limited. The quality research which does exist usually uses a sample of highly
experienced meditators rather than beginners and/or self-reports rather than
objective measures.

Over the past year, I have tried to write down what I experienced during that first
10-day course and the 10-day courses I have taken since, corroborating my experience
with research that does exist regarding meditation and the brain.

\textbf{Proviso:} S.N. Goenka, the principal teacher of Vipassana meditation,
actively dissuades students from precisely the sort of brain-centred biological
inquiry presented in this paper.

\begin{quote}
  \textbf{The brain itself is just a physical organ. As you deal with other parts of
    the body, you deal with the brain in the same way, that's all. Nothing special to
    do with the brain. But the mind is totally different. In the West, all importance
    is given to the brain as if the mind is located here. Nothing doing, it is
    everywhere. The mind is in the whole body. So give attention to the whole
    body.} \cite{goenkabrain}
\end{quote}

This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
point for those who view the brain as the quintessential physical component of mind.

\textbf{Disclaimer:} Although my primary field of research is biology, my work is far removed
from neuroscience. I have tried to simplify the research available so as to better
understand it. If there are any corrections to make or editing in terms of the
content, please feel free to get in touch with me.

\begin{center}
  \Huge{Brain Function and Anatomy}
\end{center}

Before dissecting the experience of meditation as it pertains to the brain, this
section describes the different parts of the brain that have shown up in
scientific literature as correlated to meditative practices. Neuroscience divides the
brain into the reified geography of the brain, its anatomy, and the abstract concepts
governed by brain activity, its function. Between concrete physiology and abstract
outcomes exist networks of cooperative structures which correspond to general
high-level activities of the brain.

Default Mode Network:
These are a set of structures in the mid-line of the brain (which is a evolutionarily more primitive area related to memory and emotion) which connects to structures in the cortex (which is evolutionarily the most recent part of the brain and contains the executive and higher order functions). It is the constellation of regions that light up when the brain is not engaged in any external/goal-oriented tasks (1). These regions are functionally connected and are involved in
a) The neurological basis of the self —
The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain. (Leech R, Sharp DJ (July 2013). “The role of the posterior cingulate cortex in cognition and disease”. Brain. 137 (Pt 1): 12–32.)
The PCC has been thought to mediate interactions between emotion and memory. It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes. It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).

@@

Somatosensory and Motor cortices
Another strange thing that starts happening is as your mind wandering slows down and you sit with your eyes closed, your sense of your self as a subjective agent in space and time changes, first just very slightly, as you close your eyes and meditate, it feels like you have shifted in space or turned in different directions (during these times the vmPFC and PCC which are a part of the DMN become less active), and then as the meditations get deeper and sometimes, the DMN quietened down completely for a very short duration and in addition to this, the insula is firing more rapidly, it could lead to other alternative networks of consciousness arising. It is as if losing the influence of one’s narrative leads to new insights that are usually kept from consciousness. Experiencing this even for a few seconds can lead to lasting changes even after this activity wears off. These experiences can be quite similar to being on certain kinds of drugs or even in some cases of ecstatic seizures where the abnormal activity of the anterior insula leads to heightened self awareness and feelings of bliss and lack of ambiguity. (Fabienne Picard, State of belief, subjective certainty and bliss as a product of cortical dysfunction, Cortex, Volume 49, Issue 9, 2013,
Pages 2494–2500ISSN 0010 9452, https://doi.org/10.1016/j.cortex.2013.01.006.)
So to conclude, I believe attending a ten day course literally did change the habit pattern of my mind at least for a while.
It quietens the Default Mode Network and causes the Salience Network and certain parts of the Central Executive Network to fire more rapidly (the parts that are aware of the sensations while quietening the reacting part of this network). This causes changes to a persons behavior at least for a short while and this is usually seen as soon as a person leaves the course. But repeated practice is essential for any sustained neuroplasticity since rewriting many years of habit formation requires more than a ten-day course. However, the ten day course is well designed to give the minimum observational change neccesary for people interested to keep practicing.
There have also been studies looking at the structural changes in the brain and functional changes especially to the default mode network and the salience network but these are usually measured after one has clocked over thousands of hours of practice. Therefore I am not delving into those right now.
How is this different from just concentrating on tasks hard? It is different because when concentrating on any task, your default mode network is quetened and the salience network and CEN are firing entering you into a state of flow but the reacting part of the brain is also working and working hard.
There are aspects to this technique that I have not understood or experienced that I have not delved into right now and may do once i know it better.
I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.

References

Andrews-Hanna, Jessica R. (2012–06–01). “The brain’s default network and its adaptive role in internal mentation”. The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC 3553600. PMID 21677128.
Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H. (January 2003). “Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”. Human Brain Mapping. 18 (1): 30–41.
Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan (2014–05–01). “The default network and self-generated thought: component processes, dynamic control, and clinical relevance”. Annals of the New York Academy of Sciences. 1316(1): 29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540.
Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
Isoda, M., \& Noritake, A. (2015). What makes the dorsomedial frontal cortex active during reading the mental states of others?. Neural basis of social learning, social deciding, and other-regarding preferences, 51.
Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
Menon, V; Toga, A (2015). Salience Network. Elsevier. pp. 597–611. ISBN 978–0–12–397316–0.
Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990). “The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”. Proceedings of the National Academy of Sciences of the United States of America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583.
Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002). “Dorsal anterior cingulate cortex: a role in reward-based decision making”. Proceedings of the National Academy of Sciences of the United States of America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID 11756669.
Sevinc G, Gurvit H, Spreng RN (July 2017). “Salience network engagement with the detection of morally laden information”. Social Cognitive and Affective Neuroscience. 12 (7): 1118–1127. doi:10.1093/scan/nsx035. PMID 28338944.
Jackson PL, Brunet E, Meltzoff AN, Decety J (2006). “Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”. Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID 16140345.
Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
Neuroreport. 11(7):1581–1585, May 15, 2000.
J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal, C. Morgado-Valle. The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
https://doi.org/10.1016/j.nrleng.2018.05.006.
REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
Emotion. Annals of the New York Academy of Sciences, 1121:
72–86. doi:10.1196/annals.1401.026

@@

\bibitem{goenkabrain}
  S.N. Goenka.
  \url{https://www.vridhamma.org/A-store-house-of-answers-by-Shri-S-N-Goenka}
  \textit{Answers by Mr. S. N. Goenka}
